PRISM protocol v0.3

### 2021-11-22 prism-backend/public-docs/article.tex
ATA-5717: Add protocol description
This commit adds a compact description of the protocol we use in PRISM

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\tabref}[1]{Table \ref{#1}}
\newcommand{\lstref}[1]{Listing \ref{#1}}
\newcommand{\secref}[1]{Section \ref{#1}}
\newcommand{\appref}[1]{Appendix \ref{#1}}
\newcommand{\todo}[1]{{\colorbox{red}{\bf TODO:}\textcolor{red}{#1}}}

@@

Our construction for DID management relies on an underlying blockchain. The blockchain allows us to publish transactions with sufficiently large metadata. We use the blockchain for both; data distribution along parties, and for consensus related to the order of relevant events in our protocol. All the parties participating in our protocol are running a full node that reads the metadata of the blockchain transactions. When they find a protocol event, they process it to construct their view of the system.

@@

Once the transaction is added to the blockchain with a sufficient number of confirmations $d$ (i.e. the block containing the transaction has $d$ blocks appended after it on the underlying chain), all the participants following our protocol will validate the signature of the CreateDID event. After verification, they will register the DID as \emph{published} along with all the keys posted in the initial list. The keys will be considered valid since the time associated to the event carrier transaction. This is a timestamp generated by the blockchain.

@@

Updating a DID means that the user controlling a master key associated to it, will add new or revoke existing keys to the list associated to the DID. In order to update the list of keys, the user will:
\begin{itemize}
\item Create the list of key identifiers that the user wants to revoke from the current state of his DID.
\item Create a list of keys he wants to add to the list of keys associated to his identifier. 
      This keys must be associated to a role and a fresh key identifier.
\item Create an \emph{UpdateDID} event that contains the two previously generated lists.
\item Sign the even with one of the \emph{currently associated} master keys of the DID, and publish the event inside the metadata of a transaction.
\end{itemize}

@@

In order to obtain the keys associated to a DID in short form, we simply read the state we have constructed so far by processing the CreateDID and UpdateDID events we read from the blockchain. The list of DIDs associated to the DID is the list of keys we see in our internal state. If we do not find the DID as a published one in our state, we return a \emph{DID unknown} response.

@@

Lets us define first what a credential is in the context of this document

@@

In the following steps, we assume the issuer and subjects have already established a secure communication channel.
In order to issue a batch of an arbitrary number $N$ of credentials, the issuer will:
\begin{itemize}
\item Ask to each subject the DID they would like to use for the credential they will receive.
\item The issuer creates the credentials for each subject. It uses his DID and the key id of an \emph{issuing key} to populate
      the credential corresponding fields. On each credential, it adds the corresponding subject DID as one of the claims.
\item The issuer encodes each individual credential using a base64URL encoding.
\item The issuer signs each individual encoded credential with an \emph{issuing key} associated to his publicly known DID. 
      The signing key corresponds to the key id in the credentials' claims.
\item The issuer encodes each signature, and concatenate them to their corresponding credentials using a dot (".") as separator.
      This produces strings of the form $\langle{}encoded\_credentials\rangle{}.\langle{}encoded\_signature\rangle{}$. We call these strings as \emph{signed credentials}.
\item Now, the issuer takes all the signed credentials, and computes a merkle root from them.
\item The issuer creates an \emph{IssueBatch} event that contains the merkle root, and signs the event with an \emph{issuing key} associated to his publicly known DID. 
\textbf{Note}: Today we use the same issuing key for the event signature and the individual credentials signature, but those could be different keys.
\item The issuer attaches the signed event to the metadata of a transaction and sends it to the blockchain.
\item The issuer gives to each user their corresponding signed credential along with their associated merkle inclusion proof.
\end{itemize}

@@

Note that, to issue credential batches, the issuer DID \emph{must be published} already. However, users' DIDs can remain unpublished.

@@

\subsection{Specific Sub-Protocols}
\label{ssec:subproto}

\subsubsection{End-to-End Encrypted Data Exchange via the Connector}
\label{sssec:e2eeconnector}

Assuming each participant of the system owns a (set of) DIDs, we want to let
holders receive new credentials from issuers, in a secure -- yet convenient --
manner. We describe a protocol for this purpose.  At the moment, we require
that issuer and prospective holder can perform a one-time physical interaction
with each other or, alternatively, that they have a secure channel at their
availability (although possibly cumbersome to use, hence we only require to use
it once).

In a nutshell, the proposed protocol (see \figref{fig:e2econnector}) is composed
of two phases. In the first phase, the issuer shares a QR code with the holder.
This QR code contains the issuer's public key (or, maybe, its DID), a random
value ($code$, in \figref{fig:e2econnector}), and a session identifier ($token$,
in \figref{fig:e2econnector}). This exchange takes place through a secure (i.e.,
confidential and authenticated) channel, such as physical exchange. Probably
immediately, although also possibly at a later point in time, the prospective
holder sends back her own public key (or, alternatively, her DID) authenticated
with a MAC, where we use the random value
in the QR code as secret key. The issuer uses the session identifier to locate
the corresponding session and, now that it received in an authenticated manner
the public key of the holder, can leverage hybrid encryption techniques to
establish an end-to-end encrypted session, e.g., using ECIES \cite{abr01}.
Ensuring end-to-end encryption and authentication is important at this point, as
all the information exchanged between issuer and holder is routed through the
connector: an entity or set of entities who ensure that asynchronous
communication can take place between issuer and holder. Thus, ECIES is crucial
to ensure the secrecy of the information, end-to-end. Since the holder needs to
have certainty that the data it receives comes from the issuer (and ECIES does
not authenticate the sender), we have the issuer digitally sign the data it
sends through the ECIES-secured channel. Note that we could involve the random
value shared via the QR code for this. But that would force us to re-execute the
first phase of the protocol even when the QR code is compromised after having
completed the first phase of the protocol. Digital signatures using the issuer's
keypair prevent this.

Note that we are combining signature and encryption (plus, the MAC-based
authentication natively provided by ECIES). This places us in the context put
forward in \cite{davi01}, that discusses some good practices when combining such
primitives. In a nutshell, we follow the sign-then-encrypt approach%
\footnote{What we call here ``sign-then-encrypt'' is called ``Sign \& Encrypt''
  in \cite{davi01}. However, we find
  more explicit the naming convention of \cite{kraw01}, which is more consistent
  when referring to the order in which the primitives are used -- order that, in
  these cases, is crucial.}, combined with including the public
keys of both holder and issuer into the signed (and then encrypted) data.
This intertwines the inner layer (the signature) with the outer layer (the
encryption), ensuring that any alteration on any of them will be detected.
We note that there are other options that, a priori, are equally secure -- e.g.,
encrypt-then-sign, including also the sender's public key in the encrypted data
(but that would require separately signing a (hash of) the data, to ensure
non-repudiation). Finally, it is worth noting that ECIES follows the
encrypt-then-authenticate approach, which is proven to realize secure channels
in \cite{kraw01}, as long as the underlying encryption scheme is IND-CPA secure,
and the underlying MAC function is secure -- but this is independent of the
sign-then-encrypt approach.

While the protocol is simple, and is based on common cryptographic constructions,
it is easy to get such protocols wrong, as pointed out in numerous occasions
\cite{davi01, kraw01}. Hence, we have used Tamarin%
\footnote{https://tamarin-prover.github.io/ (Last access, October 5th, 2021.)}
to model our protocol\footnote{Source code for the model available here
  \url{https://github.com/input-output-hk/atala-prism/blob/research-tamarin-connector/prism-backend/research/formalizations/connector.spthy}
  \todo{Update to final URL.}}, which is schematized in \figref{fig:e2econnector}

\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.4]{figures/connector.png}
  \caption{Sequence diagram for the E2EE protocol through the connector.}
  \label{fig:e2econnector}%
\end{figure}

The security properties that we prove are as follows (we refer to the actual
model for the formal definition):

\begin{description}
\item[Code secrecy.] Codes exchanged via a QR code are always confidential.
  Note that this is trivial, as it is exchanged via a secure channel, and
  the holder never sends it back in the clear.
\item[Code injectivity.] Every QR code that is received by a holder, allegedly
  sent by an issuer, was sent before by that issuer. Again, this is trivial due
  to the secure channel used to exchange the QR code.
\item[Holder's public key authenticity.] Every message received by an issuer
  in step 3, containing a public key for key-exchange usage, that was
  allegedly sent by a holder, was indeed previously sent by that holder.
  This is ensured through the message authentication code (MAC) that
  leverages the random QR code exchanged via the secure channel, which is
  only known to the issuer and the holder.
\item[Holder's public key injectivity.] For every session in which an
  issuer receives in step 3 a message, allegedly sent by a holder in
  step 2, containing the holder's public key, it was that holder, in
  the same session, who sent the public key in step 2.  
\item[Credential secrecy.] This property has two variants: it ensures
  secrecy for credentials sent by an honest issuer to a holder that was honest
  at the moment of sending the credential, and is not compromised afterwards;
  and it also ensures secrecy for credentials received by an honest holder,
  allegedly from an honest issuer, if both holder and issuer were honest when
  the credential is received, and they are not corrupted afterwards. Altogether
  with the credential authenticity property shown next, it ensures that only
  (honest) holder and issuer learn the contents of a credential.
\item[Credential authenticity.] All messages containing a credential, received
  by a holder at step 5, where indeed sent by an issuer and addressed to that
  same holder, as long as both issuer and holder keys are not compromised.
\item[Credential non-injectivity.] All messages containing a credential, received
  by a holder at step 5, were sent by an issuer and addressed to that holder
  at step 4. Note, however, that both messages in steps 4 and 5 don't necessarily
  pertain to the  same ``session'' (in instantiations where parallel sessions
  may occur). However, our current understanding is that this does not pose a
  risk, as in any case, the messaged is sent by the issuer, and received by the
  intended holder. For the sake of future reference, we give an illustrative
  flow of this ``non-injective'' execution, in \appref{app:non-inj-e2ee}.
\end{description}

%----------------------------------------------------------------------------------------
%	PROPERTIES
%----------------------------------------------------------------------------------------

\section{Properties we believe we have}

\paragraph{Unforgeability.} A user cannot forge issuer-generated credentials.
\paragraph{Privacy against issuers.} An issuer does not learn whether a user (who acquired a credential from the issuer) has interacted with a given verifier (assuming no cooperation between issuer and verifiers).
\paragraph{Revocability.} An issuer can revoke a previously-issued credential without users’ cooperation.
\paragraph{Protection against credential reuse.} A verifier can ensure no credential is used by different DIDs. 
\paragraph{Deferred and offline verification checks.} A verifier can verify the revocation status of a credential asynchronously without users’ participation. A verifier can also verify a credential without contacting the internet, in which case the verification result is with respect to the last time the verifier synchronized with blockchain events.
\paragraph{Resistance to key rotation.} If an issuer's issuing key is revoked, credentials already issued by that key remain valid. This means that the issuer does not need to re-issue the credentials signed by a compromised issuing key. This is possible due to the fact that key revocation and credential issuances are timestamped. Hence, if an issuing key is revoked, credentials already issued remain valid unless they are explicitly revoked.
\paragraph{Independence of issuers and verifiers.} Verifiers only need to know which issuer DIDs they trust. They do not need to contact issuers to query issuance nor revocation lists.
\paragraph{Past preservation.} Any party that verifies a credential has non-reputable information of when the credential was valid even if it has been revoked.

\section{Properties we are not offering}

\paragraph{Unlinkability.} A coalition of verifiers cannot link activities of users across verifiers.
\paragraph{Minimal disclosure.} A user only reveals what is necessary to establish a relationship with a verifier.

@@

\subsection{Scaling}

@@

The properties brought by the blockchain come at a cost. Namely, the protocol throughput (number of events that participants can attach in 
transactions per unit of time) is bounded to the blockchain throughput. In order to scale the throughput, we have reviewed Sidetree's approach. The 
idea behind Sidetree is to not publish individual events attached to blockchain transactions, but to publish a hash-link to an off-chain content 
addressable storage (CAS), like IPFS. The CAS will contain the file referenced by the hash-link, which contains a batch of many protocol events 
(allowing to transcend the size limits imposed by blockchain transactions).

@@

The first one is to add a permissioned actor (or a federation of them) that are allowed to batch protocol events and publish both on-chain hash-links,
and files in corresponding CAS. The special actor introduces a trust model that assumes that it will always reveal files. If the actor fails to 
reveal a file, the protocol participants stop processing further batches until the missing file is revealed. This is not possible in Sidetree because
any participant would be able to freeze the system by not revealing a file. The protocol would still allow users to publish events on-chain directly, 
leaving space for some decentralization for those participants that do not want to depend on the centralized batcher.

@@

\appendix

\section{Example of Non-Injective Execution of Connector E2EE Protocol}
\label{app:non-inj-e2ee}

The diagram in \figref{fig:e2econnector-noninj} depicts an scenario in which a
non-injective execution between the issuer and the holder takes place.

First, we have to take into account that such execution requires a somehow
extended setting than what we considered in \secref{sssec:e2eeconnector}. Namely,
here we assume a (certainly realistic) scenario in which any prospective holder
has several computing devices -- e.g., a laptop and a smartphone. Similarly,
the issuer instantiates several concurrent services -- depicted as
\emph{Instance 1} and \emph{Instance 2} in \figref{fig:e2econnector-noninj}.
Still, the same user operates both computing devices, which share the same
private values; and all issuer instances have access to the same database with
holders' information, and the issuer's private keys.

\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.32]{figures/connector-noninjective.png}
  \caption{Sequence diagram of a sample non-injective execution of the E2EE
    connector protocol}
  \label{fig:e2econnector-noninj}%
\end{figure}

In the diagram, the holder requests a QR code once with each device -- for this,
we envision an extended setting in which the holder can request QR codes
securely (e.g., through signed email). The smartphone communicates with Instance
2 of the issuer, while the laptop communicates with Instance 1. Eventually, the
holder decides to complete the process from her laptop. The connector forwards
the request to Instance 1, which stores it in the database. Then, it might
happen that it is Instance 2 the one that takes care of finalizing the request:
it fetches the necessary data from the shared database, and sends the result to
the holder's phone (which is the computing device that communicated with
Instance 2). This lack of correspondence between who initiated a request, and
who responds it, is what is formally referred to as non-injectivity.
Theoretically, that might perfectly happen in our proposed protocol, as there is
nothing preventing such behaviour in a cryptographic manner, as our model shows.
However, two observations are relevant:

\begin{itemize}
\item The authenticity and secrecy properties still ensure that only the holder
  and issuer have access to the data that is sent by the issuer, and that it
  indeed originates from the issuer.
\item Such behaviour, if deemed unwanted, could be easily prevented through a
  proper session handling, at the implementation level.
\end{itemize}

\bibliographystyle{plain}
\bibliography{article}

>>>>

@@

Our construction for DID management relies on an underlying blockchain. The blockchain allows us to publish transactions with sufficiently large metadata. We use the blockchain for both; data distribution along parties, and for consensus related to the order of relevant events in our protocol. All the parties participating in our protocol are running a blockchain node that reads the metadata of the blockchain transactions. When they find a protocol event, they process it to construct their view of the system.

@@

Once the transaction is added to the blockchain with a sufficient number of confirmations $d$ (i.e. the block containing the transaction has $d$ blocks appended after it on the underlying chain), all the participants following our protocol will validate the signature of the \emph{CreateDID} event. After verification, they will register the DID as \emph{published} along with all the keys posted in the initial list. The keys will be considered valid since the time associated to the event carrier transaction. This is a timestamp generated by the blockchain.

@@

Updating a DID means that the user controlling a master key associated to it, will add new or revoke existing keys to the list associated to the DID. In order to update the list of keys, the user will:
\begin{itemize}
\item Create the list of key identifiers that the user wants to revoke from the current state of his DID.
\item Create a list of keys he wants to add to the list of keys associated to his identifier. 
      This keys must be associated to a role and a fresh key identifier.
\item Create an \emph{UpdateDID} event that contains the two previously generated lists.
      The event also carries a hash of the last \emph{UpdateDID} performed over the DID (or the hash of the \emph{CreateDID} if this is the first 
      update)
\item Sign the event with one of the \emph{currently associated} master keys of the DID, and publish the event inside the metadata of a transaction.
\end{itemize}

@@

In order to obtain the keys associated to a DID in short form, we simply read the state we have constructed so far by processing the \emph{CreateDID} and \emph{UpdateDID} events we read from the blockchain. The list of keys associated to the DID, is the one we see in our internal state. If we do not find the DID as a published one in our state, we return a \emph{DID unknown} response.

@@

Lets us refine first what a credential is in the context of this document

@@

In the following steps, we assume the issuer and subjects have already established a secure communication channel. We also assume that the issuer has
a published DID that will be used to represent him in credentials.
In order to issue a batch of an arbitrary number $N$ of credentials, the issuer will:
\begin{itemize}
\item Ask to each subject the DID they would like to use for the credential they will receive.
\item The issuer creates the credentials for each subject. It uses his published DID and the key id of an \emph{issuing key} to populate
      the credential corresponding fields. On each credential, it adds the corresponding subject DID as one of the claims.
\item The issuer encodes each individual credential using a base64URL encoding.
\item The issuer signs each individual encoded credential with an \emph{issuing key} associated to his publicly known DID. 
      The signing key corresponds to the key id in the credentials' claims.
\item The issuer encodes each signature, and concatenate them to their corresponding credentials using a dot (".") as separator.
      This produces strings of the form $\langle{}encoded\_credentials\rangle{}.\langle{}encoded\_signature\rangle{}$. We call these strings asa
      \emph{signed credentials}.
\item Now, the issuer takes all the signed credentials, and computes a merkle root from them.
\item The issuer creates an \emph{IssueBatch} event that contains the merkle root and issuer DID, and signs the event with an \emph{issuing key} associated to this DID. 
\textbf{Note}: Today we use the same issuing key for the event signature and the individual credentials signature, but those could be different keys.
\item The issuer attaches the signed event to the metadata of a transaction and sends it to the blockchain.
\item The issuer gives to each subject their corresponding signed credential along with their associated merkle inclusion proof.
\end{itemize}

@@

Note that, even though the issuer DID \emph{must be published}, subjects' DIDs can remain unpublished.

@@

@@

\subsection{Increasing scalability}

@@

The properties brought by the blockchain come at a cost. Namely, the protocol throughput (number of events that participants can attach in 
transactions per unit of time) is bounded to the blockchain throughput. In order to scale the throughput, we have reviewed Sidetree's  \footnote{https://identity.foundation/sidetree/spec/} approach. The high level idea behind Sidetree is to not publish individual events attached to blockchain 
transactions, but to publish a hash-link to an off-chain content addressable storage (CAS), like IPFS. The CAS will contain the file referenced by 
the hash-link, which contains a batch of many protocol events (allowing to transcend the size limits imposed by blockchain transactions).

@@

The first one is to add a permissioned actor (or a federation of them) that are allowed to batch protocol events and publish both on-chain hash-links,
and files in corresponding CAS. The special actor introduces a trust model that assumes that it will always reveal files. If the actor fails to 
reveal a file, the protocol participants stop processing further batches until the missing file is revealed. This is not possible in Sidetree because
any participant would be able to freeze the system by not revealing a file.

This protocol variation would still allow users to publish events on-chain directly, leaving space for some decentralization for those participants that do not want to depend on the centralized batcher.

@@

At the time of this writing, we haven't implemented any of the above approaches. We had, however, implemented on-chain batching, which allows to submit multiple events in a single transaction. The events are processed in order per transaction. This has been helpful in use cases where a multiple events coming from different entities, are submitted to the same PRISM node to publish them.
